---
title: "OncoImplexus User Guide: From VCF to Interactive Reports"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    highlight: tango
---

This guide provides a comprehensive walkthrough for analyzing cancer genomes using **OncoImplexus**, covering both single-sample clinical analysis and large-scale cohort studies.

---

# 1. Quick Start with Bundled Example (DO17373)

OncoImplexus includes a bundled example dataset (DO17373) from the original ShatterSeek publication. This sample exhibits chromothripsis on chromosomes 2, 21, and X, making it ideal for testing your installation.

## Complete Example with DO17373

```{r eval=FALSE}
library(OncoImplexus)

# === Load Bundled Example Data ===
load(system.file("extdata", "DO17373.RData", package = "OncoImplexus"))

# === Create SVs Object ===
sv_obj <- SVs(
  chrom1 = SV_DO17373$chrom1,
  pos1 = SV_DO17373$start1,
  chrom2 = SV_DO17373$chrom2,
  pos2 = SV_DO17373$start2,
  strand1 = SV_DO17373$strand1,
  strand2 = SV_DO17373$strand2,
  SVtype = SV_DO17373$svclass,
  sv_id = SV_DO17373$sv_id
)

# === Create CNVsegs Object ===
cnv_obj <- CNVsegs(
  chrom = as.character(SCNA_DO17373$chromosome),
  start = SCNA_DO17373$start,
  end = SCNA_DO17373$end,
  total_cn = SCNA_DO17373$total_cn
)

# === Run Integrated Detection ===
results <- detect_chromoanagenesis(sv_obj, cnv_obj, genome = "hg19")

# === Load Gene Annotations ===
gene_granges <- readRDS(
  system.file("extdata", "hg19_genes.rds", package = "OncoImplexus")
)

# === Generate HTML Report ===
generate_interactive_report(
  result = results,
  SV.sample = sv_obj,
  CNV.sample = cnv_obj,
  output_file = "DO17373_Report.html",
  output_dir = "reports",
  sample_name = "DO17373",
  genome = "hg19",
  gene_granges = gene_granges
)
```

**Expected Output:**

- Overall classification: **Chromothripsis only**
- High confidence chromothripsis: **3 events** (chromosomes 2, 21, X)
- Possible chromoplexy: 277 chains
- Possible chromoanasynthesis: 6 regions

---

# 2. Single Sample Workflow (VCF Input)

The most common use case is analyzing a single tumor sample from VCF files produced by SV/CNV callers.

## Step 1: Loading Data from VCF

OncoImplexus supports direct VCF input. It automatically parses BND (breakend) and standard SV types.

```{r eval=FALSE}
library(OncoImplexus)

# Load SVs (supports Manta, Delly, GRIDSS, DRAGEN, Lumpy, Sniffles)
sv_data <- read_sv_vcf("tumor.sv.vcf.gz")

# Load CNVs (supports CNVkit, GATK, Control-FREEC, DRAGEN)
cnv_data <- read_cnv_vcf("tumor.cnv.vcf.gz")
```

**Note:** The VCF reader automatically detects the caller type and parses accordingly. The returned objects are already in the correct format (SVs and CNVsegs S4 classes).

## Step 2: Preparing Gene Annotations

For gene fusion prediction and driver impact analysis, you can either use the **bundled gene annotations** or load from Bioconductor databases.

### Option A: Using Bundled Annotations (Recommended)

OncoImplexus includes pre-built gene annotation files for hg19 and hg38:

```{r eval=FALSE}
# For hg38 genome
gene_granges <- readRDS(
  system.file("extdata", "hg38_genes.rds", package = "OncoImplexus")
)

# For hg19 genome
gene_granges <- readRDS(
  system.file("extdata", "hg19_genes.rds", package = "OncoImplexus")
)
```

### Option B: Using Bioconductor TxDb (Advanced)

```{r eval=FALSE}
library(TxDb.Hsapiens.UCSC.hg38.knownGene)
library(org.Hs.eg.db)

txdb <- TxDb.Hsapiens.UCSC.hg38.knownGene
genes_gr <- genes(txdb)
symbols <- mapIds(org.Hs.eg.db, keys = names(genes_gr),
                  column = "SYMBOL", keytype = "ENTREZID")
mcols(genes_gr)$symbol <- symbols
gene_granges <- genes_gr[!is.na(mcols(genes_gr)$symbol)]
```

## Step 3: Integrated Detection

Run all three mechanisms (Chromothripsis, Chromoplexy, Chromoanasynthesis) in one call:

```{r eval=FALSE}
results <- detect_chromoanagenesis(
  SV.sample = sv_data,
  CNV.sample = cnv_data,
  genome = "hg38"  # or "hg19"
)
```

## SV-only Chromoplexy Mode

If only an SV VCF is available, omit `CNV.sample`. OncoImplexus will run
chromoplexy detection only and mark the result as SV-only evidence. Chromothripsis
and chromoanasynthesis are skipped because their core evidence requires copy
number profiles.

```{r eval=FALSE}
sv_data <- read_sv_vcf("tumor.sv.vcf.gz", genome = "hg38")

results <- detect_chromoanagenesis(
  SV.sample = sv_data,
  CNV.sample = NULL,
  genome = "hg38",
  detect_chromoplexy = TRUE,
  gene_granges = gene_granges
)

results$analysis_mode
results$limitations
results$chromoplexy$summary
```

## Step 4: Generating the Interactive HTML Report

This is the final deliverable, producing a clinical-grade dashboard with visualizations.

```{r eval=FALSE}
generate_interactive_report(
  result = results,
  SV.sample = sv_data,
  CNV.sample = cnv_data,
  output_file = "Patient_X_Report.html",
  output_dir = "reports",
  sample_name = "Patient X (Tumor)",
  genome = "hg38",
  gene_granges = gene_granges
)
```

**Important Parameters:**

| Parameter | Description |
|-----------|-------------|
| `result` | The output from `detect_chromoanagenesis()` |
| `SV.sample` | The SVs object from `read_sv_vcf()` |
| `CNV.sample` | The CNVsegs object from `read_cnv_vcf()` |
| `output_file` | Output HTML filename (just the filename, not path) |
| `output_dir` | Directory to save the report |
| `sample_name` | Display name in the report |
| `genome` | Reference genome ("hg19" or "hg38") |
| `gene_granges` | Gene annotation GRanges object |

## Complete VCF Example Script

Here's a complete working example using VCF files:

```{r eval=FALSE}
library(OncoImplexus)

# === Configuration ===
sample_id <- "H435"
sv_vcf <- "path/to/H435.sv.vcf.gz"
cnv_vcf <- "path/to/H435.cnv.vcf.gz"
output_dir <- "reports"
genome <- "hg38"

# === Step 1: Load Data ===
sv_data <- read_sv_vcf(sv_vcf)
cnv_data <- read_cnv_vcf(cnv_vcf)

# === Step 2: Load Gene Annotations ===
gene_granges <- readRDS(
  system.file("extdata", "hg38_genes.rds", package = "OncoImplexus")
)

# === Step 3: Run Analysis ===
results <- detect_chromoanagenesis(sv_data, cnv_data, genome = genome)

# === Step 4: Generate Report ===
generate_interactive_report(
  result = results,
  SV.sample = sv_data,
  CNV.sample = cnv_data,
  output_file = paste0(sample_id, "_Report.html"),
  output_dir = output_dir,
  sample_name = sample_id,
  genome = genome,
  gene_granges = gene_granges
)
```

---

# 3. Using Data Frames (Advanced)

If you have SV/CNV data in custom formats (e.g., from non-standard callers or downstream processing), you can create the S4 objects manually.

## Creating SVs Object from Data Frame

```{r eval=FALSE}
# Your SV data frame must have these columns
sv_df <- data.frame(
  chrom1 = c("1", "2", "1"),
  pos1 = c(1000000, 5000000, 8000000),
  chrom2 = c("1", "3", "1"),
  pos2 = c(2000000, 8000000, 9000000),
  strand1 = c("+", "-", "+"),
  strand2 = c("-", "+", "+"),
  SVtype = c("DEL", "TRA", "h2hINV"),
  sv_id = c("SV1", "SV2", "SV3")
)

# Create SVs object
sv_obj <- SVs(
  chrom1 = sv_df$chrom1,
  pos1 = sv_df$pos1,
  chrom2 = sv_df$chrom2,
  pos2 = sv_df$pos2,
  strand1 = sv_df$strand1,
  strand2 = sv_df$strand2,
  SVtype = sv_df$SVtype,
  sv_id = sv_df$sv_id
)
```

**SV Types:**

| SVtype | Description | Strand Pattern |
|--------|-------------|----------------|
| DEL | Deletion-like | +/- |
| DUP | Duplication-like | -/+ |
| h2hINV | Head-to-head inversion | +/+ |
| t2tINV | Tail-to-tail inversion | -/- |
| TRA | Translocation | varies |

## Creating CNVsegs Object from Data Frame

```{r eval=FALSE}
# Your CNV data frame must have these columns
cnv_df <- data.frame(
  chr = c("1", "1", "2"),
  start = c(1, 5000000, 1),
  end = c(4999999, 10000000, 50000000),
  cn = c(2, 1, 3)
)

# Create CNVsegs object
cnv_obj <- CNVsegs(
  chrom = cnv_df$chr,
  start = cnv_df$start,
  end = cnv_df$end,
  total_cn = cnv_df$cn
)
```

**Note:** Chromosome names must be in Ensembl notation (1-22, X, Y) without "chr" prefix.

---

# 4. Cohort Analysis Workflow

For research studies involving multiple patients, follow the batch processing pipeline.

## Step 1: Batch Processing

Process a list of samples using a loop:

```{r eval=FALSE}
library(OncoImplexus)

sample_list <- c("SP1001", "SP1002", "SP1003")
data_dir <- "vcf_files"
output_dir <- "cohort_results"

# Create output directory
dir.create(output_dir, showWarnings = FALSE, recursive = TRUE)

# Load gene annotations once
gene_granges <- readRDS(
  system.file("extdata", "hg19_genes.rds", package = "OncoImplexus")
)

for (sid in sample_list) {
  message("Processing: ", sid)

  # Load data
  sv_data <- read_sv_vcf(file.path(data_dir, paste0(sid, ".sv.vcf.gz")))
  cnv_data <- read_cnv_vcf(file.path(data_dir, paste0(sid, ".cnv.vcf.gz")))

  # Run detection
  results <- detect_chromoanagenesis(sv_data, cnv_data, genome = "hg19")

  # Save results
  saveRDS(results, file = file.path(output_dir, paste0(sid, ".rds")))

  # Generate individual reports
  generate_interactive_report(
    result = results,
    SV.sample = sv_data,
    CNV.sample = cnv_data,
    output_file = paste0(sid, "_Report.html"),
    output_dir = output_dir,
    sample_name = sid,
    genome = "hg19",
    gene_granges = gene_granges
  )
}
```

## Step 2: Aggregate & Summarize

Use the built-in summary tool to create a master CSV table:

```{r eval=FALSE}
summary_table <- summarize_cohort_results(
  results_dir = "cohort_results",
  output_csv = "Study_Summary.csv"
)

# View summary metrics
head(summary_table)
```

**Master Summary Table Columns:**

| Column | Description |
|--------|-------------|
| `sample_id` | Sample name (from .rds filename) |
| `classification` | Integrated diagnosis (e.g., "Multiple mechanisms") |
| `ct_hc` | High Confidence chromothripsis events |
| `cp_likely` | Likely chromoplexy chains |
| `cs_likely` | Likely chromoanasynthesis regions |
| `n_fusions` | Predicted gene fusions |

**Example Output:**

| sample_id | classification | ct_hc | cp_likely | cs_likely | n_fusions |
|-----------|----------------|-------|-----------|-----------|-----------|
| PATIENT_01 | Chromothripsis only | 2 | 0 | 0 | 8 |
| PATIENT_02 | Multiple mechanisms | 1 | 42 | 0 | 15 |
| PATIENT_03 | No chromoanagenesis | 0 | 0 | 0 | 0 |

---

# 5. Interpreting the Results

## High vs. Low Confidence (Chromothripsis)

- **High Confidence**: Passes the "Fragment Joins Test" (p > 0.05) and has strong CN oscillations (>=7 segments). This is the "gold standard" for chromothripsis.
- **Low Confidence**: Clustered SVs with fewer CN oscillations (4-6). May represent early or degraded events.

## Likely vs. Possible (Chromoplexy)

- **Likely**: Chains with 3+ chromosomes and high CN stability.
- **Possible**: Chains that are borderline or involve only 2 chromosomes.

## Likely vs. Possible (Chromoanasynthesis)

- **Likely**: Regions with clear copy number gradients AND supporting SV topology evidence.
- **Possible**: Regions with partial evidence (gradient only or topology only).

---

# 6. Troubleshooting FAQ

**Q: Why does my report say 'N/A' for DNA Repair?**

A: Ensure the corresponding BSgenome package is installed:
```{r eval=FALSE}
BiocManager::install("BSgenome.Hsapiens.UCSC.hg38")  # for hg38
BiocManager::install("BSgenome.Hsapiens.UCSC.hg19")  # for hg19
```

**Q: I get 'SV.sample must be a SVs object' error.**

A: Make sure you're using an SVs object, not a data frame. If reading from VCF, `read_sv_vcf()` returns the correct object. If using custom data, create the object manually (see Section 3).

**Q: I get 'cannot open connection' errors during report rendering.**

A: This usually happens if the data is too large. OncoImplexus v1.0.0+ uses direct environment injection to solve this. Ensure you are using the latest version.

**Q: My SV caller is not recognized.**

A: The VCF reader auto-detects most callers (Manta, Delly, GRIDSS, DRAGEN, Lumpy, Sniffles). If your caller is not recognized, create the SVs object manually from a data frame (see Section 3).

**Q: How do I suppress browser tabs opening during report generation?**

A: This is fixed in v1.0.0+. The report template now includes options to prevent automatic browser opening.

**Q: Chromosome name errors?**

A: OncoImplexus uses Ensembl chromosome notation (1-22, X, Y) without "chr" prefix. The VCF readers handle this automatically, but if using custom data, ensure chromosomes are named correctly.

**Q: How do I analyze hg19 vs hg38 samples?**

A: Simply specify the correct genome parameter and use the matching gene annotations:
```{r eval=FALSE}
# For hg38
gene_gr_38 <- readRDS(system.file("extdata", "hg38_genes.rds",
                                   package = "OncoImplexus"))
results <- detect_chromoanagenesis(sv_data, cnv_data, genome = "hg38")

# For hg19
gene_gr_19 <- readRDS(system.file("extdata", "hg19_genes.rds",
                                   package = "OncoImplexus"))
results <- detect_chromoanagenesis(sv_data, cnv_data, genome = "hg19")
```
