---
title: "OncoImplexus Cohort Analysis Report"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    theme: united
    toc: true
    toc_float: true
    code_folding: hide
params:
  data_file: "path/to/data.rds"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.width = 12, fig.height = 8)
library(OncoImplexus)
library(ggplot2)
library(DT)
library(reshape2)

# Load data passed from R function
data_list <- readRDS(params$data_file)
cohort <- data_list$cohort
res_list <- cohort@results
n_samples <- length(cohort@sample_ids)
```

# 1. Cohort Overview

This report summarizes chromoanagenesis patterns across a cohort of **`r n_samples`** samples.

```{r agg_stats}
# Robust aggregation
agg_list <- list()
for (sid in names(res_list)) {
  r <- res_list[[sid]]
  s <- r$integrated_summary
  
  if (is.null(s)) {
    # Create empty row if missing
    agg_list[[sid]] <- data.frame(
      sample_id = sid, CT_high = 0, CT_low = 0, CP_likely = 0, CP_possible = 0,
      CS_likely = 0, CS_possible = 0, Overall = "Not Analyzed", stringsAsFactors = FALSE
    )
  } else {
    # Use helper to ensure single values
    get_val <- function(v, default=0) {
      if (is.null(v) || length(v) == 0 || is.na(v)) return(default)
      return(v[1])
    }
    
    agg_list[[sid]] <- data.frame(
      sample_id = sid,
      CT_high = get_val(s$chromothripsis_high_confidence),
      CT_low = get_val(s$chromothripsis_low_confidence),
      CP_likely = get_val(s$chromoplexy_likely),
      CP_possible = get_val(s$chromoplexy_possible),
      CS_likely = get_val(s$chromoanasynthesis_likely),
      CS_possible = get_val(s$chromoanasynthesis_possible),
      Overall = get_val(s$overall_classification, "No events"),
      stringsAsFactors = FALSE
    )
  }
}
agg_df <- do.call(rbind, agg_list)

# Frequency Calculation
freq_stats <- data.frame(
  Mechanism = c("Chromothripsis", "Chromoplexy", "Chromoanasynthesis"),
  Likely_High = c(
    sum(agg_df$CT_high > 0, na.rm=T),
    sum(agg_df$CP_likely > 0, na.rm=T),
    sum(agg_df$CS_likely > 0, na.rm=T)
  ),
  Possible_Low = c(
    sum(agg_df$CT_low > 0, na.rm=T),
    sum(agg_df$CP_possible > 0, na.rm=T),
    sum(agg_df$CS_possible > 0, na.rm=T)
  )
)
```

## 1.1 Global Prevalence
Proportion of samples exhibiting each chromoanagenesis mechanism.

```{r prevalence_plot}
df_plot <- melt(freq_stats[, 1:3], id.vars = "Mechanism")
ggplot(df_plot, aes(x = Mechanism, y = value, fill = variable)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme_minimal() +
  labs(y = "Number of Samples", fill = "Confidence", title = "Mechanism Prevalence across Cohort") +
  scale_fill_manual(values = c("Likely_High" = "#d32f2f", "Possible_Low" = "#ffcdd2"))
```

# 2. Co-occurrence & Overlap

Analysis of how different mechanisms occur together in the same tumor.

```{r co_occurrence}
binary_mat <- data.frame(
  Chromothripsis = ifelse(agg_df$CT_high > 0 | agg_df$CT_low > 0, 1, 0),
  Chromoplexy = ifelse(agg_df$CP_likely > 0 | agg_df$CP_possible > 0, 1, 0),
  Chromoanasynthesis = ifelse(agg_df$CS_likely > 0 | agg_df$CS_possible > 0, 1, 0)
)

if (requireNamespace("UpSetR", quietly = TRUE)) {
  UpSetR::upset(binary_mat, sets = c("Chromothripsis", "Chromoplexy", "Chromoanasynthesis"), 
        order.by = "freq", main.bar.color = "#455a64")
} else {
  # Simple Heatmap Fallback
  co_occ <- t(as.matrix(binary_mat)) %*% as.matrix(binary_mat)
  diag(co_occ) <- colSums(binary_mat)
  co_df <- melt(co_occ)
  ggplot(co_df, aes(Var1, Var2, fill = value)) +
    geom_tile() +
    geom_text(aes(label = value), color = "white") +
    scale_fill_gradient(low = "white", high = "#455a64") +
    theme_minimal() +
    labs(title = "Mechanism Co-occurrence Matrix", x="", y="", fill="Count")
}
```

# 3. Targeted Driver Genes

Most frequently impacted cancer driver genes across the cohort.

```{r gene_hotspots}
all_driver_hits <- lapply(res_list, function(r) {
  if(!is.null(r$gene_annotations)) return(r$gene_annotations$driver_hits)
  if(!is.null(r$annotation)) return(r$annotation$driver_hits)
  return(NULL)
})
gene_df <- do.call(rbind, Filter(Negate(is.null), all_driver_hits))

if (!is.null(gene_df) && nrow(gene_df) > 0) {
  gene_counts <- as.data.frame(table(gene_df$Gene, gene_df$Mechanism))
  colnames(gene_counts) <- c("Gene", "Mechanism", "Frequency")
  
  top_genes <- aggregate(Frequency ~ Gene, data = gene_counts, sum)
  top_genes <- top_genes$Gene[order(top_genes$Frequency, decreasing = TRUE)][1:min(20, nrow(top_genes))]
  
  ggplot(gene_counts[gene_counts$Gene %in% top_genes, ], aes(x = reorder(Gene, Frequency), y = Frequency, fill = Mechanism)) +
    geom_bar(stat = "identity") +
    coord_flip() +
    theme_minimal() +
    labs(x = "Gene Symbol", title = "Top Impacted Driver Genes")
} else {
  cat("No high-confidence driver gene hits found in this cohort.")
}
```

# 4. Sample Detail Catalog

Detailed classification for every sample in the cohort.

```{r detail_table}
datatable(agg_df, 
          filter = "top", 
          options = list(pageLength = 15, scrollX = TRUE),
          caption = "Cohort Sample Summary")
```

---
Generated by **OncoImplexus** | `r Sys.time()`
