---
title: "OncoImplexus Survival Analysis Report"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    theme: united
    toc: true
    toc_float: true
    code_folding: hide
params:
  data_file: "path/to/data.rds"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.width = 10, fig.height = 6)
library(OncoImplexus)
library(ggplot2)
data_list <- readRDS(params$data_file)
surv_result <- data_list$survival_result
datatable_or_table <- function(x, caption = NULL) {
  if (requireNamespace("DT", quietly = TRUE)) {
    DT::datatable(x, filter = "top", options = list(pageLength = 15, scrollX = TRUE), caption = caption)
  } else {
    knitr::kable(x, caption = caption)
  }
}
```

# 1. Analysis Overview

## `r data_list$title`

This report tests whether OncoImplexus-derived chromoanagenesis features are
associated with survival outcomes.

```{r overview}
overview <- data.frame(
  metric = c("Samples with usable survival data", "Features tested", "Cox rows", "KM/log-rank rows"),
  value = c(
    nrow(surv_result$analysis_data),
    length(surv_result$feature_cols),
    nrow(surv_result$cox_table),
    nrow(surv_result$km_table)
  ),
  stringsAsFactors = FALSE
)
datatable_or_table(overview, "Survival analysis overview")
```

# 2. Cox Regression

Univariable Cox regression is run for every selected feature. If covariates were
provided, multivariable Cox results are also shown.

```{r cox_table}
if (!is.null(surv_result$cox_table) && nrow(surv_result$cox_table) > 0) {
  show_cols <- intersect(c("feature", "model", "n", "n_events", "hazard_ratio",
                           "conf_low", "conf_high", "p_value", "fdr", "covariates"),
                         colnames(surv_result$cox_table))
  datatable_or_table(surv_result$cox_table[, show_cols, drop = FALSE], "Cox regression results")
} else {
  cat("No Cox regression results were generated.")
}
```

```{r forest_plot}
if (!is.null(surv_result$cox_table) && nrow(surv_result$cox_table) > 0) {
  print(plot_survival_forest(surv_result, model = "multivariable", top_n = 20))
}
```

# 3. Kaplan-Meier / Log-Rank Analysis

Binary features are compared as absent/present. Continuous features are split by
the configured cutpoint, usually the median.

```{r km_table}
if (!is.null(surv_result$km_table) && nrow(surv_result$km_table) > 0) {
  show_cols <- intersect(c("feature", "n", "n_events", "cutpoint", "groups",
                           "chisq", "p_value", "fdr"),
                         colnames(surv_result$km_table))
  datatable_or_table(surv_result$km_table[, show_cols, drop = FALSE], "KM/log-rank results")
} else {
  cat("No KM/log-rank results were generated.")
}
```

```{r top_km_plots, results='asis', fig.width=9, fig.height=5}
if (!is.null(surv_result$km_table) && nrow(surv_result$km_table) > 0) {
  top_features <- head(surv_result$km_table$feature[order(surv_result$km_table$p_value)], 5)
  for (feature in top_features) {
    cat("\n\n## Kaplan-Meier:", feature, "\n\n")
    print(plot_survival_km(surv_result, feature))
  }
}
```

# 4. Feature Matrix

```{r feature_matrix}
if (!is.null(surv_result$features) && nrow(surv_result$features) > 0) {
  datatable_or_table(surv_result$features, "OncoImplexus survival feature matrix")
}
```

---
Generated by **OncoImplexus** | `r Sys.time()`
